Short-time expansion of Heisenberg operators in open collective quantum spin systems

### 2020-02-04 soc_squeezing/collective_paper/collective_spin.tex
pushing final, published versions of the collective spin paper

The truncation strategy in Eq.~\eqref{eq:weight_truncation} has a few
limitations:
\begin{enumerate*}
\item simulating a system of differential equations for a large number
  of operators can be time-consuming,
\item the weight measure $w$ may need to be chosen carefully, as the
  optimal measure is generally system-dependent, and
\item simulation results can only be trusted up to the time at which
  the initial value of operators $\hat\S_{\v m}$ with weight
  $w\p{\v m}\ge W$ have a non-negligible contribution to expectation
  values of interest.\label{pt:limitation}
\end{enumerate*}
The last limitation in particular unavoidably applies in some form to
any method tracking only a subset of all relevant operators.  We
therefore devise an alternate truncation strategy built around
limitation \ref*{pt:limitation}.

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \bk{\hat\O\p{t}}
  = \bk{e^{t\check T} \hat \O\p{0}}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator $\check T^k$ are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \bk{\hat\O\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Note that when computing an
expectation value $\bk{\hat\O\p{t}}$, the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, cuts both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}$ and the number of
matrix elements $T_{\v m\v n}$ that we may need to explicitly compute
roughly in half.

@@

Three primary considerations limit the maximum time $t$ to which we
can accurately compute a correlator $\bk{\hat S_{\v n}\p{t}}$ using
the TST expansion.  First, maintaining accuracy at larger times $t$
requires going to higher orders $M$ in the TST expansion.  An
order-$M$ TST expansion of the correlator $\bk{\hat S_{\v n}\p{t}}$
can involve a significant fraction of operators $\hat S_{\v m}$ with
weight $\abs{\v m}\lesssim M$, which implies the need to compute
$O\p{M^3}$ initial-time expectation values $\bk{\hat S_{\v m}}$ and
$O\p{M^4}$ matrix elements $T^k_{\v m\v n}$.  In practice, with a
straightforward implementation of the TST expansion we find that these
requirements generally restrict $M\lesssim 50$ -- $70$ with $8$ --
$50$ gigabytes of random access memory (RAM).  Second, individual
terms at high orders of the TST expansion in Eq.~\eqref{eq:TST} can
grow excessively large, greatly amplifying any numerical errors and
thereby spoiling cancellations that are necessary to arrive at a
physical value of a correlator, i.e.~with
$\abs{\bk{\hat S_{\v n}\p{t}}}\lesssim S^{\abs{\v n}}$.  Finally, the
TST expansion is essentially perturbative in the time $t$, which
implies that its validity as a formal expansion eventually breaks
down.  Precisely characterizing the implications of these last two
considerations for the TST expansion requires additional analysis that
we defer to future work.  An investigation of connections between the
TST expansion and past work related to the Mori
formalism\cite{mori1965continuedfraction, annett1994recursive}, for
example, might answer questions about the breakdown and convergence of
the TST expansion.  As we show from benchmarks of the TST expansion in
Section \ref{sec:squeezing}, however, a detailed understanding of
breakdown is not necessary to diagnose the breakdown time
$t_{\t{break}}^{(M)}$ beyond which the TST expansion yields inaccurate
results.  Empirically, we find that going beyond order $M\approx35$
yields no significant gains in all cases considered in this work.

@@

In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters for diagnosing time-crystalline
phases of matter\cite{tucker2018shattered}.  Squared commutators such
as that in Eq.~\eqref{eq:four_point}, meanwhile, are commonly examined
for signatures of quantum chaos and information
scrambling\cite{maldacena2016bound, swingle2018unscrambling,
  garcia-mata2018chaos}.  In typical scenarios, such squared
commutators initially vanish by construction through to a choice of
spatially separated operators.  Collective spin systems, however, have
no intrinsic notion of locality or spatial separation.  In our case,
therefore, with the choice of initial state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$ we merely have
$D\p{0}\sim1/N$.

@@

Figure \ref{fig:multi_time} shows the behavior of $C\p{t}$ and
$D\p{t}$ for $N=10^4$ spins, initially in the state $\ket\X$, evolving
under the squeezing Hamiltonians in
Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both with and without spontaneous
decay, excitation, and dephasing of individual spins at rates
$\gamma_-=\gamma_+=\gamma_\z=100\chi$.  In the case of unitary
evolution under OAT, we find that to an excellent approximation
$\abs{C\p{t}}$ takes the functional form
$f\p{t}=f\p{0}+aN\chi t+\p{bN\chi t}^2$ with $a\sim b\sim 1$, and with
a virtually perfect fit $D\p{t}=D\p{0}+\p{\sp{N+1}\chi t}^2$.  For
unitary evolution under TAT and TNT, we find that to an excellent
approximation both $\abs{C\p{t}}$ and $D\p{t}$ take the functional
form $f\p{t}=f\p{0}+a\sp{\exp\p{bN\chi t}-1}$ with $a\sim b\sim 1$.
As may be expected, the growth of $C\p{t}$ and $D\p{t}$ is generally
suppressed by decoherence.  Figure \ref{fig:multi_time} serves as an
example for the type of behavior that is accessible at short times
with the TST expansion.  These examples are straightforward extend to
equilibrium settings and spin-boson systems.

@@

\begin{acknowledgments}
  We acknowledge helpful discussions with Robert Lewis-Swan, Kris
  Tucker, and Colin Kennedy; as well as some technical contributions
  from Diego Barberena.  This work is supported by the Air Force
  Office of Scientific Research (AFOSR) grant FA9550-18-1-0319; the
  AFOSR Multidisciplinary University Research Initiative (MURI) grant;
  the Defense Advanced Research Projects Agency (DARPA) and Army
  Research Office (ARO) grant W911NF-16-1-0576 and W911NF-19-1-0210;
  the National Science Foundation (NSF) grant PHY-1820885; JILA-NSF
  grant PFC-173400; and the National Institute of Standards and
  Technology (NIST).
\end{acknowledgments}

@@

We now need rearrange the operators in $C_{mn;\mu}$ into a standard

@@

Here we discuss the details of emulating particle loss with $O(1/N)$
error, where $N$ is the initial number of particles in a system that
we wish to describe with a spin model.  Starting with the full algebra
of creation and annihilation operators (whether bosonic or fermionic)
in a system, spin models are typically implemented by identifying a
subalgebra of relevant ``spin'' operators that satisfy appropriate
commutation relations.  Two-state particles on a lattice, for example,
are described by annihilation operators $c_{js}$ indexed by a lattice
site $j\in\mathbb{Z}$ and an internal state index $s\in\set{\up,\dn}$,
enabling the straightforward construction of spin operators
\begin{align}
  \sigma_\x^{(j)} \equiv c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  &&
  \sigma_\y^{(j)} \equiv - i c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  &&
  \sigma_\z^{(j)}
  \equiv c_{j,\up}^\dag c_{j,\up} - c_{j,\dn}^\dag c_{j,\dn},
  &&
  \1^{(j)} \equiv c_{j,\up}^\dag c_{j,\up} + c_{j,\dn}^\dag c_{j,\dn},
\end{align}
which satisfy same commutation relations as the standard Pauli
operators.  These spin operators can be more compactly defined in the
form
\begin{align}
  \sigma_\alpha^{(j)} \equiv \sum_{r,s\in\set{\up,\dn}}
  c_{jr}^\dag \bk{r|\sigma_\alpha|s} c_{js},
\end{align}
where $\sigma_\alpha$ for $\alpha\in\set{\x,\y,\z,\1}$ is a Pauli
operator, with $\sigma_\1\equiv\1$; and $\bk{r|\sigma_\alpha|s}$
denotes a matrix element of $\sigma_\alpha$.  This construction
exemplifies how the set of jump operators
$\J_{\t{loss}}^{\t{bare}}\equiv\set{c_{js}}$ that generate particle
loss cannot be constructed from spin operators, which are generally
bilinear in particle creation/annihilation operators.  When working on
the level of a spin model, therefore, we can at best only {\it
  emulate} the effect of particle loss by some indirect means.

@@

Here we provide supplementary evidence of our finding in Section
\ref{sec:squeezing} that the TNT model can produce more squeezing than
the OAT or TAT models in the presence of strong decoherence.  To this
end, Figure \ref{fig:sqz_dec_scaling} shows the minimal squeezing
parameter $\xi^2_{\t{min}}$ achievable with $N=100$ spins through the
OAT, TAT, and TNT models as a function of the rate $\gamma_0$ at which
individual spins undergo spontaneous decay, excitation, and dephasing.
These results were computed with quantum trajectory simulations, with
$10^3$ trajectories per data point.  While the OAT and TAT models
produce more squeezing than TNT model with weak decoherence, this
squeezing falls off faster with an increasing decoherence rate
$\gamma_0$.  The relative robustness of TNT is in part a consequence
of the fact that TNT initially generates squeezing at a faster rate
than OAT or TAT, thereby allowing it to produce more squeezing before
the degrading effects of decoherence kick in.

>>>>

The truncation strategy in Eq.~\eqref{eq:weight_truncation} has a few
limitations:
\begin{enumerate*}
\item simulating a system of differential equations for a large number
  of operators can be time-consuming,
\item the weight measure $w$ may need to be chosen carefully, as the
  optimal measure is generally system-dependent, and
\item simulation results can only be trusted up to the time at which
  the initial values of operators $\hat\S_{\v m}$ with weight
  $w\p{\v m}\ge W$ have a non-negligible contribution to expectation
  values of interest.\label{pt:limitation}
\end{enumerate*}
The last limitation in particular unavoidably applies in some form to
any method tracking only a subset of all relevant operators.  We
therefore devise an alternate truncation strategy built around
limitation \ref*{pt:limitation}.

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \bk{\hat\O\p{t}}
  = \bk{e^{t\check T} \hat \O\p{0}}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator $\check T^k$ are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and zero otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \bk{\hat\O\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
We refer to Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Note that when computing an
expectation value $\bk{\hat\O\p{t}}$, the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, cuts both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}$ and the number of
matrix elements $T_{\v m\v n}$ that we may need to explicitly compute
roughly in half.

@@

Three primary considerations limit the maximum time $t$ to which we
can accurately compute a correlator $\bk{\hat S_{\v n}\p{t}}$ using
the TST expansion.  First, maintaining accuracy at larger times $t$
requires going to higher orders $M$ in the TST expansion.  An
order-$M$ TST expansion of the correlator $\bk{\hat S_{\v n}\p{t}}$
can involve a significant fraction of operators $\hat S_{\v m}$ with
weight $\abs{\v m}\lesssim M$, which implies the need to compute
$O\p{M^3}$ initial-time expectation values $\bk{\hat S_{\v m}}$ and
$O\p{M^4}$ matrix elements $T^k_{\v m\v n}$.  In practice, with a
straightforward implementation of the TST expansion we find that these
requirements generally restrict $M\lesssim 50$ -- $70$ with $8$ --
$50$ gigabytes of random access memory (RAM).  Second, individual
terms at high orders of the TST expansion in Eq.~\eqref{eq:TST} can
grow excessively large, greatly amplifying any numerical errors and
thereby spoiling cancellations that are necessary to arrive at a
physical value of a correlator, i.e.~with
$\abs{\bk{\hat S_{\v n}\p{t}}}\lesssim S^{\abs{\v n}}$ (where
$S\equiv N/2$).  Finally, the TST expansion is essentially
perturbative in the time $t$, which implies that its validity as a
formal expansion eventually breaks down.  Precisely characterizing the
implications of these last two considerations for the TST expansion
requires additional analysis that we defer to future work.  An
investigation of connections between the TST expansion and past work
related to the Mori formalism\cite{mori1965continuedfraction,
  annett1994recursive}, for example, might answer questions about the
breakdown and convergence of the TST expansion.  As we show from
benchmarks of the TST expansion in Section \ref{sec:squeezing},
however, a detailed understanding of breakdown is not necessary to
diagnose the breakdown time $t_{\t{break}}^{(M)}$ beyond which the TST
expansion yields inaccurate results.  Empirically, we find that going
beyond order $M\approx35$ yields no significant gains in all cases
considered in this work.

@@

In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters for diagnosing time-crystalline
phases of matter\cite{tucker2018shattered}.  Squared commutators such
as that in Eq.~\eqref{eq:four_point}, meanwhile, are commonly examined
for signatures of quantum chaos and information
scrambling\cite{maldacena2016bound, swingle2018unscrambling,
  garcia-mata2018chaos}.  In typical scenarios, such squared
commutators initially vanish by construction through a choice of
spatially separated operators.  Collective spin systems, however, have
no intrinsic notion of locality or spatial separation.  In our case,
therefore, with the choice of initial state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$ we merely have
$D\p{0}\sim1/N$.

@@

Figure \ref{fig:multi_time} shows the behavior of $C\p{t}$ and
$D\p{t}$ for $N=10^4$ spins, initially in the state $\ket\X$, evolving
under the squeezing Hamiltonians in
Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both with and without spontaneous
decay, excitation, and dephasing of individual spins at rates
$\gamma_-=\gamma_+=\gamma_\z=100\chi$.  In the case of unitary
evolution under OAT, we find that to an excellent approximation
$\abs{C\p{t}}$ takes the functional form
$f\p{t}=f\p{0}+aN\chi t+\p{bN\chi t}^2$ with $a\sim b\sim 1$, and with
a virtually perfect fit $D\p{t}=D\p{0}+\p{\sp{N+1}\chi t}^2$.  For
unitary evolution under TAT and TNT, we find that to an excellent
approximation both $\abs{C\p{t}}$ and $D\p{t}$ take the functional
form $f\p{t}=f\p{0}+a\sp{\exp\p{bN\chi t}-1}$ with $a\sim b\sim 1$.
As may be expected, the growth of $C\p{t}$ and $D\p{t}$ is generally
suppressed by decoherence.  Figure \ref{fig:multi_time} serves as an
example for the type of behavior that is accessible at short times
with the TST expansion.  These examples are straightforward to extend
to equilibrium settings and spin-boson systems.

@@

\begin{acknowledgments}
  We acknowledge helpful discussions with Robert Lewis-Swan, Kris
  Tucker, and Colin Kennedy; as well as some technical contributions
  from Diego Barberena.  This work is supported by the Air Force
  Office of Scientific Research (AFOSR) grant FA9550-18-1-0319; the
  Defense Advanced Research Projects Agency (DARPA) and Army Research
  Office (ARO) grant W911NF-16-1-0576 and W911NF-19-1-0210; the
  National Science Foundation (NSF) grant PHY-1820885; JILA-NSF grant
  PFC-1734006; and the National Institute of Standards and Technology
  (NIST).
\end{acknowledgments}

@@

We now need to rearrange the operators in $C_{mn;\mu}$ into a standard

@@

Here we discuss the details of emulating particle loss with $O(1/N)$
error, where $N$ is the initial number of particles in a system that
we wish to describe with a spin model.  Starting with the full algebra
of creation and annihilation operators (whether bosonic or fermionic)
in a system, spin models are typically implemented by identifying a
subalgebra of relevant ``spin'' operators that satisfy appropriate
commutation relations.  Two-state particles on a lattice, for example,
are described by annihilation operators $c_{js}$ indexed by a lattice
site $j\in\mathbb{Z}$ and an internal state index $s\in\set{\up,\dn}$,
enabling the straightforward construction of spin operators
\begin{align}
  \sigma_\x^{(j)} \equiv c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  &&
  \sigma_\y^{(j)} \equiv - i c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  &&
  \sigma_\z^{(j)}
  \equiv c_{j,\up}^\dag c_{j,\up} - c_{j,\dn}^\dag c_{j,\dn},
  &&
  \1^{(j)} \equiv c_{j,\up}^\dag c_{j,\up} + c_{j,\dn}^\dag c_{j,\dn},
\end{align}
which satisfy the same commutation relations as the standard Pauli
operators.  These spin operators can be more compactly defined in the
form
\begin{align}
  \sigma_\alpha^{(j)} \equiv \sum_{r,s\in\set{\up,\dn}}
  c_{jr}^\dag \bk{r|\sigma_\alpha|s} c_{js},
\end{align}
where $\sigma_\alpha$ for $\alpha\in\set{\x,\y,\z,\1}$ is a Pauli
operator, with $\sigma_\1\equiv\1$; and $\bk{r|\sigma_\alpha|s}$
denotes a matrix element of $\sigma_\alpha$.  This construction
exemplifies how the set of jump operators
$\J_{\t{loss}}^{\t{bare}}\equiv\set{c_{js}}$ that generate particle
loss cannot be constructed from spin operators, which are generally
bilinear in particle creation or annihilation operators.  When working
on the level of a spin model, therefore, we can at best only {\it
  emulate} the effect of particle loss by some indirect means.

@@

Here we provide supplementary evidence of our finding in Section
\ref{sec:squeezing} that the TNT model can produce more squeezing than
the OAT or TAT models in the presence of strong decoherence.  To this
end, Figure \ref{fig:sqz_dec_scaling} shows the minimal squeezing
parameter $\xi^2_{\t{min}}$ achievable with $N=100$ spins through the
OAT, TAT, and TNT models as a function of the rate $\gamma_0$ at which
individual spins undergo spontaneous decay, excitation, and dephasing.
These results were computed with quantum trajectory simulations, with
$10^3$ trajectories per data point.  While the OAT and TAT models
produce more squeezing than the TNT model with weak decoherence, this
squeezing falls off faster with an increasing decoherence rate
$\gamma_0$.  The relative robustness of TNT is in part a consequence
of the fact that TNT initially generates squeezing at a faster rate
than OAT or TAT, thereby allowing it to produce more squeezing before
the degrading effects of decoherence kick in.
